### 2011-04-04 paper.tex
original

\documentclass[10pt]{IEEEtran}
\usepackage{setspace}
\usepackage{psfig}
\usepackage{boxedminipage}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{cite}
\usepackage{times}
\usepackage[pdftex]{graphicx}
\usepackage{comment}
\usepackage{subfigure}
\usepackage{wrapfig}
% \usepackage{amsthm}
% \usepackage{enumerate}
\usepackage{multirow}

\begin{document}

\title{Bandwidth-Aware Reconfigurable Memory Design with Hybrid Memory
  Technologies}

% \author{\\
% \{\}@psu.edu\vspace{-10pt}
% }

\maketitle

\begin{abstract}
  In future chip-multiprocessor~(CMP) design, memory bandwidth is a potential
  bottleneck to system performance. Emerging non-volatile memory technologies,
  such as spin-torque-transfer memory~(STT-RAM), resistive memory~(RRAM), and
  embedded DRAM~(eDRAM), are promising solutions as on-chip cache and memory for
  CMPs. In this paper, we propose a bandwidth-aware reconfigurable memory
  hierarchy design with hybrid memory technologies. We demonstrate a method to
  dynamically reconfigure the number of shared memory levels and capacity of
  each level based on statistical prediction. With a set of both multithreaded
  and multiprogrammed applications, we evaluate system performance obtained with
  our method. The experimental results show that the proposed reconfigurable
  hybrid memory design improves the overall system throughput by \% compared to
  pure SRAM memory design. In addition, our design leads to \% throughput
  increase and \% dynamic power reduction compared to hybrid but fixed memory
  hierarchy design.

\end{abstract}

% ------------------------------------------------------------------------------

\section{Introduction}

One potential bottleneck for chip-multiprocessor~(CMP) performance scaling is
the widening gap between the bandwidth demand created by processor cores and the
limited access bandwidth provided by off-chip main memory [15][16][6][20]. Such
limitation greatly affects the parallelism of memory demanding applications,
i.e., applications with a large working set, by consuming additional cycles on
off-chip memory access. In addition, even moderate memory demand applications
will be affected as the number of cores scales up [17]. The continuing shrink of
transistor density not only leads to increasingly powerful processors with a
large number of cores, but also puts much bandwidth pressure to off-chip main
memory. However, the bandwidth to the off-chip main memory does not improve much
compared to the processor core scaling. Potentially, these issues make the
applications running on the computing systems be bandwidth-bound at main memory.

While a number techniques can be found in today's systems and research work to
alleviate such bandwidth bottleneck, few can improve the system throughput in a
power and cost efficient manner. High performance computing machines such as
NVIDIA's Tesla~\cite{nvidia:tesla} rely on very high main memory bandwidths
(provided by GDDR memories) to feed the requirement from the processor. However,
GDDR memories run at higher clock rates and are more power hungry than
conventional DRAM modules. It is undesirable for either general purpose or high
performance computing systems to improve their computing performance by
sacrifacing power efficiency. With the emerging 3D chip technology, caches can
be stacked on top of processor cores to provide high memory
bandwidth[12][21][22]. However, ...

Caching has long been employed as the most effective approach to reduce memory
access latency. Adding an extra level of cache, however, can also help alleviate
the increasing bandwidth pressure of off-chip memory. In [dac.1] the authors
discuss extensively the problem of limited pin bandwidth to multiprocessor
systems. They have focused on program performance in multiprocessor systems and
make detailed decomposition of program execution times. They conclude that even
more complex on-chip cache structures would prove to be costeffective, with the
limited pin bandwidth severely restricting performance increases. In [dac.2] the
authors have carefully studied the requirements for on-chip cache structures
along with all sorts of optimization techniques that come along with scaling of
processor core numbers. Their study shows that cache size need to grow much
faster than processor core numbers to compensate for the limited off-chip
bandwidth. Because of that, the near future processors need to allocate a huge
percentage of chip area for caches, which means much less core counts than
expected. The study also shows that effective bandwidth optimization techniques
can help reduce cache size requirement and thus help scaling processor cores.
However, the effectiveness of caching in the mid/large scale multiprocessor
systems could be highly suboptimal due to significant contention across the
parallel tasks. The more bandwidth requirement by a workload can be satisfied by
the last level cache~(LLC), the less bandwidth will be required from the
off-chip main memory. Our memory technology pre-exploration shows that the
bandwidth provided by a memory decreases with the increase of its capacity.
Based on these observations, adding more levels of cache can potentially fill in
the bandwidth gap.

On-chip caches with fast random access, high storage density, and non-volatility
become possible due to the emergence of various new non-volatile memory~(NVM)
technologies, such as spin-torque-transfer memory~(STT-RAM), phase-change
memory~(PCRAM), and resistive memory~(RRAM). These emerging memory technologies
are believed to be promising solution as on-chip
caches~\cite{Sun:2009:MRAM-L2-CMP}. In addition to the benefit of
non-volatility, we demonstrate that these memory technologies provide higher
bandwidth than SRAM at a high capacity. Consequently, it is beneficial to use
NVM as lower level caches when large capacity.

This paper presents a bandwidth-aware memory hierarchy design to enhance system
performance of CMPs. Our goal is to enhance the memory system from the bandwidth
point of view, by leveraging these emerging memory technologies in devising a
hybrid cache hierarchy. The contributions we present in this paper include:

\begin{itemize}
\item Bandwidth-aware hybrid on-chip memory hierarchy design.
\item Hardware support that enables prediction of cache performance on the
  different sized caches.
\item OS scheduler support to make use of the prediction capability and
  appropriately schedule applications on to core with suitable cache capacity.

\end{itemize}

% -----------------------------------------------------------------------------

\section{Related Work}

The bandwidth problem of multicore processors has drawn much attention recently.
Yu \emph{et al.} proposed a LLC partitioning algorithm to minimize bandwidth
requirement to off-chip main memory~\cite{Yu:2010:OCMB}. While most cache
partitioning techniques focus on cache miss rates, our work takes a different
approach in which tasks¡¯ memory bandwidth requirements are taken into account
when identifying a cache partitioning for multi-programmed and/or multithreaded
workloads. Cache resources are allocated with the objective that the overall
system bandwidth requirement is minimized for the target workload. The key
insight is that cache miss-rate information may severely misrepresent the actual
bandwidth demand of the task, which ultimately determines the overall system
performance and power consumption.

A large body of recent research focuses on exploring new memory technologies to
trade off between latency, bandwidth, and cost. Wu et al. explored various
memory technologies - SRAM, eDRAM [23], magnetic RAM (MRAM) [12] and PCM to best
construct L3 caches in terms of performance and power [24]. The eDRAM and MRAM
technologies are evaluated as last-level caches by stacking each memory on top
of the processor die. Both studies mainly focus on reducing the latency gap
between L2 cache and external memory and do not examine the bandwidth. <nvm
cache research: most for latency, not bandwidth-aware opt.>

Application behavior prediction is a critical component of reconfigurable
architectures. Zhou \emph{et al.} monitored memory access patterns and estimated
memory behavior of workloads for energy efficient memory allocation
[predictor.26]. Duesterwald et al. [10] describe different statistical and table
based predictors for within- and across-metric predictions of performance
monitoring information. They show that the table-based predictor generally
outperforms the other predictors they tested. Sarikaya et al. [19] describe an
optimal prediction technique based on a predictive least squares minimization.
Sarikaya \emph{et al.} showed the benefit of statistical metric modeling for
tracking varying pattern history lengths and modeling long term
patterns~\cite{Sarikaya:2011:RWBP}. However, ...

% -----------------------------------------------------------------------------

\section{Background}

\noindent{}{\bf{STT-RAM}} STT-RAM uses the magnetic property of the material and
uses Magnetic Tunnel Junction (MTJ) as its binary storage. As shown in
Fig.~\ref{fig:mram_cell}, MTJ contains two ferromagnetic layers and one tunnel
barrier layer. The direction of one ferromagnetic layer is fixed, which is
called the reference layer, while the direction of the other one can be changed
by passing a driving current, which is called the free layer. The relative
magnetization direction of two ferromagnetic layers determines the resistance of
MTJ. If two ferromagnetic layers have the same directions, the resistance of MTJ
is low, indicating a ``0'' state; if two layers have different directions, the
resistance of MTJ is high, indicating a ``1'' state.

\begin{figure}[t]
\centering
\includegraphics[width=2.5in]{figures/mram_cell}
\caption{Demonstration of a MRAM cell. (a) Structural view. (b) Schematic view.}
\label{fig:mram_cell}
\end{figure}

% \noindent{}{\bf{PCRAM}} PCRAM uses chalcogenide-based material to storage
% informations. The chalcogenide-based materials in recent PCRAM research are
% usually alloys of germanium, antimony, and tellurium (GeSbTe, or GST), which can
% be switched between a crystalline phase (SET or ``1'' state) and an amorphous
% phase (RESET or ``0'' state) with the application of heat. The crystalline phase
% shows high optical reflectivity and low electrical resistivity, while the
% amorphous phase is characterized by low reflectivity and high resistivity.

% \begin{figure}[t]
% \centering
% \includegraphics[width=2.5in]{figures/GST}
% \caption{The schematic view of a PCRAM cell with MOSFET selector transistor (BL=Bitline, WL=Wordline, SL=Sourceline}
% \label{fig:GST}
% \end{figure}

\noindent{}{\bf{RRAM}}

\begin{figure}[t]
\centering
\includegraphics[width=2.5in]{figures/memristor}
\caption{The conceptual view of the structure of memristor cells.}
\label{fig:memristor}
\end{figure}

Memristor, a portmanteau of ``memory resistor'', is a generalized resistance
that maintains a functional relationship between the time integrals of current
and voltage. Memristor was first theoretically predicted by Chua in
1971~\cite{memristor:chua} as the fourth fundamental circuit element from the
completeness of relations between the four basic circuit variables, namely,
current, voltage, charge, and flux-linkage. The first memristor practical
demonstration was presented by Williams \emph{et al.} in
2008~\cite{memristor:missing}. Fig.~\ref{fig:memristor} shows a conceptual view
of the memristor structure~\cite{memristor:missing}. The top electrode and
bottom electrode are two metal nanowires on platinum, and the thin titanium
dioxide film is sandwiched by the electrodes.

% -----------------------------------------------------------------------------
\section{Memory Technology Exploration}
Many modeling tools have been developed during the last decade to enable
system-level design exploration for SRAM- or DRAM-based cache and memory design.
For example, CACTI~\cite{CACTI} is a tool that has been widely used in the
computer architecture community to estimate the speed, power, and area of SRAM
and DRAM caches. In addition, CACTI has also been extended to evaluate the
performance, power, and area for STT-RAM~\cite{CACTI:DAC08:Dong},
PCRAM~\cite{CACTI:GLSVLSI08:Mangalagiri,CACTI:PCRAMsim}, and NAND
flash~\cite{CACTI:DATE10:Mohan}. However, as CACTI is originally designed to
model SRAM-based cache, some of its fundamental assumptions do not match the
actual NVM circuit implementation, and thereby these CACTI-like estimation tools
do not model the NVM array organization in the exact way that the chip is
fabricated. In this section, we use \emph{NVSim}, a circuit-level model for NVM
performance, energy, and area estimation, which supports various NVM
technologies including STT-RAM, PCRAM, RRAM, and conventional NAND flash.

\begin{figure*}[t]
\centering
\includegraphics[width=5in]{figures/RAM_bandwidth}
\caption{Read and write bandwidths provided by different memory technologies.
(a) Read bandwidth provided by different memory technologes. (b) Write bandwidth
provided by different memory technologies.}
\label{fig:memory-bw}
\end{figure*}

First of all, we estimate the read and write bandwidths that can be provided by
different memory technologies. Figure~\ref{fig:memory-bw} shows the results,
with both x- and y-values in \emph{log} scale. The figure illustrates both the
provided read and write bandwidth as a function of memory capacity. Each of the
memory technologies actually provide nearly the same read bandwidths, as is
shown in figure~\ref{fig:memory-bw}(a). On the other hand, a straight forward
observation from figure~\ref{fig:memory-bw}(b) is that the write bandwidth
varies among different memory technologies. The shape of the SRAM write
bandwidth curve is very similar to the read bandwidth curve. The write bandwidth
curves of the other three memory technologies appear to be very different. The
reason is that write latencies of the three non-volatile memories are much
higher than read latencies. Another observation from
figure~\ref{fig:memory-bw}(b) is that the curves cross to each other at
different locations.

\begin{figure*}[t]
\centering
\includegraphics[width=5in]{figures/RAM_latency}
\caption{Latency of different memory technologies. (a) Read Latency. (b) Write Latency.}
\label{fig:memory-latency}
\includegraphics[width=5in]{figures/RAM_energy}
\caption{Dynamic energy consumption with the provided bandwidths of different
  memory technologies. (a) Dynamic Energy with 20\% write.  (b) Dynamic Energy with 50\% write. }
\label{fig:memory-energy}
\end{figure*}

Being aware that the read and write latencies of NVM are asymmetric, we consider
the read latency at first. In NVsim, we divide the entire cache read latency
into the following components:
\begin{enumerate}
\item H-tree input delay
\item Decoder + word-line delay
\item Bit-line delay + Sense Amplifier delay
\item Comparator Delay (for tag part only)
\item H-tree output delay
\end{enumerate}
H-tree latencies 1) and 5) are mainly determined by the RC delay of global
wires, which is positive proportional to the area of memory macro. Sensing delay
4) is related to the read noise margin of memory cell that is affected by off/on
resistance ratio. Figure~\ref{fig:memory-latency} (a) illustrates the read
latency of different memories. We can see that sensing delay dominates the read
latency of NVM at small capacity so that PCRAM (with the largest resistance
window) is faster than RRAM and MRAM (with the smallest resistance window).
While H-tree delay unveils at large capacity so that RRAM (with the smallest
cell size) becomes faster than PRAM and MRAM (with the biggest cell size). The
read latency of SRAM bank will increase rapidly after 128MB due to large area.

Write latency of NVM are almost dominated by the write pulse width. In this work
we assume 10ns, 20ns, 100ns for MRAM, RRAM and PCRAM. While write latency of
SRAM and eDRAM is a function of capacity, as similar to the read latency. The
results in ~\ref{fig:memory-latency} (b) indicates that NVM is suitable for
memory with large capacity.

Figure~\ref{fig:memory-energy} has demonstrated the dynamic energy of different
memory technologies when 20\% and 50\% write access are assumed. eDRAM will be
better than SRAM in terms of dynamic energy after 16MB and this is verified by
IBM Power7 L3 cache. The cross-point between NVM and SRAM/eDRAM is postponed for
50\% write than 20\% write.

Based on these observations, we would like to design a bandwidth-aware hybrid
memory hierarchy, which always provides the high memory bandwidth with the given
capacity. For example,
\begin{itemize}
\item eDRAM has better latency and enegry than SRAM when capacity is larger than
  16MB.
\item MRAM is more competitive to SRAM and eDRAM when capacity is larger than
  128MB.
\item PCRAM has serious endurance issue and is targeted as main memory
  replacement.
\item RRAM might fit into cache hierarchy as last level cache replacement when
  there multiple levels of cache for applications with very few writes.
\end{itemize}

\section{Design Method}

\section{Experiments}

Based on the parameters of different cache configurations collected from our
modified version of CACTI~\cite{CACTI}, we evaluate both pure SRAM-based and
hybrid cache hierarchy designs. In this section, we show how bandwidth-aware
memory hierarchy design method help with improving the system performance.

\subsection{Experimental Setup}

We use Simics~\cite{Magnusson:2002:simics-orig-paper} as the simulator in our
experiments. It is configured to model an eight-core CMP. Each core is in-order, and
is similar to UltraSPARC III architecture. The frequency of each core is set to
be 1GHz. Table~\ref{table:cmp-config} lists the detailed parameters of the
baseline. Our baseline contains 8 private L1 instruction and data caches
respectively. Since we only evaluate shared cache hierarchies, each of L1 cache
is fixed to be SRAM-based, and of 64KB capacity. As regard to lower level
caches, we evaluate two cases, pure SRAM-based and hybrid caches with various
memory technologies, including SRAM, MRAM, RRAM, and eDRAM. By evaluating both
cases, we would like to find out optimal cache design that leads to the peak
performance, i.e., the number of cache levels that is reqired, memory technology
used for each level, and capacity of each level.

\begin{table}[t]
  \centering
  \caption{Baseline CMP configuration.}
  \begin{tabular}{l|c}
    \hline\hline
    \multicolumn{2}{l}{\textbf{Core}}\\
    \hline
    No. of cores & 8\\
    \hline
    Frequency & 1GHz\\
    \hline
    Core architecture & in-order, 14-stage pipeline\\
    \hline
    \multicolumn{2}{l}{\textbf{Memory}}\\
    \hline
    \multirow{2}{*}{Private caches} & L1-I/D caches: SRAM, 8 x 64KB,\\
                                    & 64B line, 2-way, write-through \\                                    
    \hline
    \multirow{2}{*}{Shared caches} & Case 1: Pure SRAM, 64B line,\\
                                   & 8-way, write-back\\
                                   & Case 2: Hybrid~(SRAM, MRAM,\\
                                   & RRAM, eDRAM), 64B line,\\
                                   & 8-way, write-back\\
    \hline
    Main memory & 4GB\\
    \hline\hline
  \end{tabular}
  \label{table:cmp-config}
\end{table}

The benchmarks are selected from PARSEC benchmark suite~\cite{PARSEC:2008} with
multithreaded programs, which focus on emerging workloads that are designed to
be representative of next-generation shared-memory programs for CMPs. Since the
performance of different memory technologies are closely related to read and
write intensities, we selected some workloads that vary in the average numbers
of L2 cache read per thousand instructions~(RPKI) and write per thousand
instructions~(WPKI), which are listed in
Table~\ref{table:benchmarks}.

\begin{table}[t]
  \centering
  \caption{Characteristics of selected benchmarks.}
  \begin{tabular}{l|c|c}
    \hline\hline    
    \textbf{Benchmarks} & \textbf{RPKI} & \textbf{WPKI}\\
    \hline
    blackscholes & 22.8 & 61.7\\
    \hline
    bodytrack & 5.4 & 139.5\\
    \hline
    canneal & 5.4 & 29.3\\
    \hline
    facesim & 6.0 & 102.4\\ 
    \hline
    ferret & 5.7 & 173.4\\
    \hline
    fluidanimate & 2.6 & 70.1\\
    \hline
    streamcluster & 17.3 & 16.9\\
    \hline
    swaptions & 2.6 & 121.4\\
    \hline\hline
  \end{tabular}
  \label{table:benchmarks}
\end{table}

\subsection{Results}

We evaluate various possible configurations with shared caches by simulation,
i.e., with possible numbers of levels, memory technologies, and cache capacities
for each cache level. We consider SRAM and eDRAM as the possible memory
technologies to implement L2 and L3 caches. Both MRAM and RRAM have higher write
latency than SRAM. Furthermore, the endurance of RRAM is too low to be used as
lower level caches. Consequently, these two memory technologies are only
considered to be used as the last level cache.

In the first set of experiments, we evaluate the system performance with
two-level shared caches. Figure~\ref{fig:thput-l2l3} shows the system throughput
with all the benchmarks. It is illustrated that implementing hybrid cache with
eDRAM as the L3 cache helps with performance among most of the benchmarks. Such
performance improvement is more than 10\% in Figure~\ref{fig:thput-l2l3}(a). It
is indicated by our memory technology exploration that eDRAM shows more latency
benefits with larger capacities. As a result, a larger eDRAM-based L3 cache
leads to more performance improvement to the pure SRAM implementation, as
illustrated by Figure~\ref{fig:thput-l2l3}(b). However, hybrid cache does not
always improve the performance, as shown in Figure~\ref{fig:benchmark-l2l3}.
With the benchmark \emph{canneal}, hybrid cache configurations outperform the
pure SRAM implementation among overall range of various capacities. With the
benchmark \emph{ferret}, however, pure SRAM implementation results in higher
performance than hybrid cache implementation with the same capacity
configuration.

\begin{figure*}[t]
\centering
\includegraphics[width=2.5in]{figures/thput-l2l3-1}\hspace{0.2in}
\includegraphics[width=2.5in]{figures/thput-l2l3-2}\\
\hspace{0.02in}
\makebox[2.5in][l]{\bf (a)}
\makebox[2.5in][l]{\bf (b)}
\caption{Performance comparison with two-level caches among various benchmarks.
  (a) The L3 cache capacity is 16MB. (b) The L3 cache capacity is 32MB.}
\label{fig:thput-l2l3}
\end{figure*}

\begin{figure*}[t]
\centering
\includegraphics[width=2.5in]{figures/canneal-l2l3}\hspace{0.2in}
\includegraphics[width=2.5in]{figures/ferret-l2l3}\\
\hspace{0.02in}
\makebox[2.5in][l]{\bf (a)}
\makebox[2.5in][l]{\bf (b)}
\caption{Performance comparison with two-level caches among various cache
  capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
\label{fig:benchmark-l2l3}
\end{figure*}

In the second set of experiments, we evaluate the system performance with
three-level shared caches. When evaluating hybrid cache configurations, we
consider implementing the last level cache by MRAM and RRAM memory technologies,
since the data transaction intensity is relatively low in the last level cache.
Figure~\ref{fig:thput-l2l3l4} shows the results.
Figure~\ref{fig:thput-l2l3l4}(a) illustrated that the performance more than 15\%
higher on average with a last level cache implemented by MRAM than pure SRAM
implementations. More performance improvement is obtained by increasing the last
level cache capacity, as shown in Figure~\ref{fig:thput-l2l3l4}(b).
Figure~\ref{fig:benchmark-l2l3l4} compares performance with pure SRAM and hybrid
cache implementations with different capacities. In this case, hybrid cache
implementation leads to higher performance in both benchmarks. The results of
the two sets of experiments are illustrated together in
Figure~\ref{fig:benchmark-overall}. An interesting observation is that hybrid
cache implementation shows high performance improvement with one of the
applications \emph{canneal}, whereas as little improvement with the other
\emph{ferret}. The reason is that write latency of both MRAM and RRAM is higher
than SRAM when capacity is less than 1GB. Performance of applications with large
number of writes, such as \emph{ferret} is therefore not benefit from hybrid
cache design.

\begin{figure*}[t]
\centering
\includegraphics[width=2.5in]{figures/thput-l2l3l4-1}\hspace{0.2in}
\includegraphics[width=2.5in]{figures/thput-l2l3l4-2}\\
\hspace{0.02in}
\makebox[2.5in][l]{\bf (a)}
\makebox[2.5in][l]{\bf (b)}

\caption{Performance comparison with three-level caches among various
  benchmarks. (a) The L3 cache capacity is 16MB, and the L4 cache capacity is
  512MB. (b) The L3 cache capacity is 32MB, and the L4 cache capacity is 1GB.}
\label{fig:thput-l2l3l4}
\end{figure*}

\begin{figure*}[t]
\centering
\includegraphics[width=2.5in]{figures/canneal-l2l3l4}\hspace{0.2in}
\includegraphics[width=2.5in]{figures/ferret-l2l3l4}\\
\hspace{0.02in}
\makebox[2.5in][l]{\bf (a)}
\makebox[2.5in][l]{\bf (b)}
\caption{Performance comparison with three-level caches among various cache
  capacity. (a) The canneal benchmark. (b) The ferret benchmark..}
\label{fig:benchmark-l2l3l4}
\end{figure*}

\begin{figure*}[t]
\centering
\includegraphics[width=2.5in]{figures/canneal}\hspace{0.2in}
\includegraphics[width=2.5in]{figures/ferret}\\
\hspace{0.02in}
\makebox[2.5in][l]{\bf (a)}
\makebox[2.5in][l]{\bf (b)}

\caption{Performance comparison with both two-level and three-level caches among
  various cache capacity. (a) The canneal benchmark. (b) The ferret benchmark.}
\label{fig:benchmark-overall}
\end{figure*}

Finally, we evaluate all the possible cache hierarchy configurations with
exhaustive simulations. The optimal cache configurations of each benchmark is
shown in Table~\ref{table:results}. Since the characteristics vary among the
benchmarks, the optimal cache configurations are different among each of
benchmarks. Applications with high write intensities, such as \emph{bodytrack},
\emph{ferret}, and \emph{swaptions} tend to favor SRAM-based caches. Other
applications benefit more from hybrid cache implementation in terms of
performance.

\begin{table*}[t]
  \centering
  \caption{Optimal cache configurations for each benchmark.}
  \vspace{0.1in}
  \begin{tabular}{l|c|c|c}
    \hline\hline
    \textbf{Benchmarks} & \textbf{L2} & \textbf{L3} & \textbf{L4}\\
    \hline
    blackscholes & SRAM, 8MB & eDRAM, 32MB & RRAM, 1GB\\
    \hline
    bodytrack & SRAM, 8MB & eDRAM, 128MB & -\\
    \hline
    canneal & SRAM, 8MB & eDRAM, 512MB & MRAM, 1GB\\
    \hline
    facesim & SRAM, 8MB & eDRAM, 32MB & MRAM, 1GB\\
    \hline
    ferret & SRAM, 8MB & eDRAM, 32MB & -\\
    \hline
    fluidanimate & SRAM, 8MB & eDRAM, 1GB & -\\
    \hline
    streamcluster & SRAM, 16MB & eDRAM, 128MB & MRAM, 512MB\\
    \hline
    swaptions & SRAM, 8MB & SRAM, 32MB & SRAM, 1GB\\
    \hline\hline
  \end{tabular}
  \label{table:results}
\end{table*}

\section{Conclusion and Future Work}
In this project, we explore performance and energy characteristics of various
emerging memory technolgies. Based on our exploration, we evaluate the shared
cache hierarchy design of CMPs optimized for performance by filling the off-chip
memory bandwidth gap. According to our evaluation, SRAM-based caches leads to
reasonable performance with write-intensive applications. Hybrid cache design
help with performance with other types of benchmarks.

Furthermore, a variety of continuing studies is remained to be explored related
to bandwidth-aware memory hierarchy design. First of all, we only examine cache
hierarchy design currently. It is necessary to extend our exploration to the
overall memory hierarchy design. On the other hand, we do not consider energy
efficiency in this project. We would like to put some power constraints to the
memory hierarchy, and examine the energy efficiency of various memory hierarchy
configurations.

\bibliographystyle{IEEEtran}
\bibliography{./bib/ref,./bib/reram,./bib/cacti}

\end{document}

### 2011-04-04 paper.tex

@@
 is low, indicating a ``0'' state; if two layers have different directions, the
-resistance of MTJ is high, indicating a ``1'' state.
+resistance of MTJ is high, indicating a ``1'' state. \vspace{0.1in}
 
@@
 
-\noindent{}{\bf{RRAM}}
+\noindent{}{\bf{RRAM}} Memristor, a portmanteau of ``memory resistor'', is a
+generalized resistance that maintains a functional relationship between the time
+integrals of current and voltage. Memristor was first theoretically predicted by
+Chua in 1971~\cite{memristor:chua} as the fourth fundamental circuit element
+from the completeness of relations between the four basic circuit variables,
+namely, current, voltage, charge, and flux-linkage. The first memristor
+practical demonstration was presented by Williams \emph{et al.} in
+2008~\cite{memristor:missing}. Fig.~\ref{fig:memristor} shows a conceptual view
+of the memristor structure~\cite{memristor:missing}. The top electrode and
+bottom electrode are two metal nanowires on platinum, and the thin titanium
+dioxide film is sandwiched by the electrodes.
 
@@
 
-Memristor, a portmanteau of ``memory resistor'', is a generalized resistance
-that maintains a functional relationship between the time integrals of current
-and voltage. Memristor was first theoretically predicted by Chua in
-1971~\cite{memristor:chua} as the fourth fundamental circuit element from the
-completeness of relations between the four basic circuit variables, namely,
-current, voltage, charge, and flux-linkage. The first memristor practical
-demonstration was presented by Williams \emph{et al.} in
-2008~\cite{memristor:missing}. Fig.~\ref{fig:memristor} shows a conceptual view
-of the memristor structure~\cite{memristor:missing}. The top electrode and
-bottom electrode are two metal nanowires on platinum, and the thin titanium
-dioxide film is sandwiched by the electrodes.
+\noindent{}{\bf{eDRAM??}}
 
@@
 \section{Design Method}
+
+\subsection{Memory Hierarchy Design}
+
+\subsection{Reconfiguration}
+
+\subsection{Workload Behavior Prediction}
 

### 2011-04-04 trunk/paper.tex
haha

--- 2011-04-04 paper.tex
@@
 
+% -------------------------------------------------------------------------
 \section{Design Method}
 
+\subsection{Memory Hierarchy Design}
+
+\subsection{Reconfiguration}
+
+\subsection{Workload Behavior Prediction}
+
+% -------------------------------------------------------------------------
 \section{Experiments}
@@
 
+% --------------------------------------------------------------------------
 \section{Conclusion and Future Work}

### 2011-04-04 paper.tex

@@
   such as spin-torque-transfer memory~(STT-RAM), resistive memory~(RRAM), and
-  embedded DRAM~(eDRAM), are promising solutions as on-chip cache and memory for
-  CMPs. In this paper, we propose a bandwidth-aware reconfigurable memory
-  hierarchy design with hybrid memory technologies. We demonstrate a method to
-  dynamically reconfigure the number of shared memory levels and capacity of
+  embedded DRAM~(eDRAM), are promising solutions as on-chip memories for CMPs.
+  In this paper, we propose a bandwidth-aware reconfigurable on-chip
+  memory~(BROM) design with hybrid memory technologies. We demonstrate a method
+  to dynamically reconfigure the number of shared memory levels and capacity of
   each level based on statistical prediction. With a set of both multithreaded
@@
 bandwidth demand of the task, which ultimately determines the overall system
-performance and power consumption.
+performance and power consumption. <the moguls paper, the figure 1, however design for
+induvidual application, i.e., with different application can generate different
+designs.>
 
@@
 construct L3 caches in terms of performance and power [24]. The eDRAM and MRAM
-technologies are evaluated as last-level caches by stacking each memory on top
-of the processor die. Both studies mainly focus on reducing the latency gap
-between L2 cache and external memory and do not examine the bandwidth. <nvm
-cache research: most for latency, not bandwidth-aware opt.>
+technologies are evaluated as LLCs by stacking each memory on top of the
+processor die. Both studies mainly focused on reducing the latency gap between L2
+cache~(LLC) and external memory and do not examine the bandwidth. <nvm cache
+research: most for latency, not bandwidth opt.>
 
@@
 
+% -------------------------------------------------------------------------
 \section{Design Method}
+
+While many design methods envolved with new memory technologies endeavor to
+reduce the memory access latency of going off-chip, this work employs the hybrid
+on-chip memory and reconfiguration to minimize off-chip bandwidth requirement.
+For each level of memory, we select the memory technology that has the highest
+provided bandwidth within given capacity. On top of such hardware hierarchy
+design, we apply dynamic memory system reconfiguration that awares of the
+bandwidth demand of a specific set of applications.
+
+Based on our memory technology exploration, we now provide an overview of our
+bandwidth-aware architecture (Section~\ref{sec:hierarchy}) , and then discuss the details on the hardware
+support (Section 3.2) and OS support (Section 3.3) required for achieving high
+performance on this architecture.
 
@@
 
-\subsection{Workload Behavior Prediction}
+tailoring mode
+reuse mode
 
+\subsection{Prediction Engine}
+
+statistical prediction
+performance opt
+
+% -------------------------------------------------------------------------
 \section{Experiments}

### 2011-04-04 paper.tex

@@
 
-While many design methods envolved with new memory technologies endeavor to
-reduce the memory access latency of going off-chip, this work employs the hybrid
-on-chip memory and reconfiguration to minimize off-chip bandwidth requirement.
-For each level of memory, we select the memory technology that has the highest
-provided bandwidth within given capacity. On top of such hardware hierarchy
-design, we apply dynamic memory system reconfiguration that awares of the
-bandwidth demand of a specific set of applications.
-
-Based on our memory technology exploration, we now provide an overview of our
-bandwidth-aware architecture (Section~\ref{sec:hierarchy}) , and then discuss the details on the hardware
-support (Section 3.2) and OS support (Section 3.3) required for achieving high
-performance on this architecture.
+Based on our memory technology exploration, we now provide an overview of BROM
+design. While many design methods envolved with new memory technologies endeavor
+to reduce the off-chip memory access latency, this work focuses on minimize
+off-chip bandwidth requirement by employing hybrid on-chip memory and
+reconfiguration. For each level of memory, we select the memory technology that
+has the highest provided bandwidth within given capacity. On top of such
+hardware hierarchy design, we apply dynamic memory system reconfiguration with
+the bandwidth demand of a specific set of applications as the major metric. One
+desirable characteristic of BROM is that it provides prioritized access to the
+residual memory capacity available at runtime instead of sharing it among all
+data flows. By having control over the residual capacity, we can better utilize
+the limited resources to enhance system performance. The hardware
+reconfiguration is further facilitated by software prediction engine. Rather
+than conventional last value predictors and table-based predictors, we employ a
+probablity-based predictor that can achieve higher accuracy with small
+performance overhead.
 

### 2011-04-04 trunk/paper.tex

@@
 
+Based on our memory technology exploration, we now provide an overview of BROM
+design. While many design methods envolved with new memory technologies endeavor
+to reduce the off-chip memory access latency, this work focuses on minimize
+off-chip bandwidth requirement by employing hybrid on-chip memory and
+reconfiguration. For each level of memory, we select the memory technology that
+has the highest provided bandwidth within given capacity. On top of such
+hardware hierarchy design, we apply dynamic memory system reconfiguration with
+the bandwidth demand of a specific set of applications as the major metric. One
+desirable characteristic of BROM is that it provides prioritized access to the
+residual memory capacity available at runtime instead of sharing it among all
+data flows. By having control over the residual capacity, we can better utilize
+the limited resources to enhance system performance. The hardware
+reconfiguration is further facilitated by software prediction engine. Rather
+than conventional last value predictors and table-based predictors, we employ a
+probablity-based predictor that can achieve higher accuracy with small
+performance overhead. The rest of this section describes the BROM design, which
+is composed of on-chip memory hierarchy hardware design, the reconfiguration
+mechanism, and the predictor engine.
+
 \subsection{Memory Hierarchy Design}
+
+hardware design: num of levels; memory technology of each level; size of each level
+
+reconfiguration support: how reconfigurable (Li -> Li+1 (Li+0.5); hardware
+modification to support reconfiguration; what to do with reconfigured capacity
 
@@
 
-\subsection{Workload Behavior Prediction}
+decide size of each level si:
+
+1) theoritical curves
+
+2) bound:
+
+- dbwi < pbwi -> si <= s1
+
+- max throughput under power budget? -> si >= s2
+
+- si <= hardware given max si
+
+- si > si-1
+
+reconfigure the unused capacity:
+
+- if (> si): becomes si+1
+
+- else: unused (or becomes lut)
+
+\subsection{Prediction Engine}
+
+statistical prediction
+
+hierarchical prediction
+
+performance opt: lut long enough but not too long (num of entries, bits per entry)
 

### 2011-04-05 trunk/paper.tex

@@
   In this paper, we propose a bandwidth-aware reconfigurable on-chip
-  memory~(BROM) design with hybrid memory technologies. We demonstrate a method
+  memory~(BAROM) design with hybrid memory technologies. We demonstrate a method
   to dynamically reconfigure the number of shared memory levels and capacity of
@@
 
-Based on our memory technology exploration, we now provide an overview of BROM
+Based on our memory technology exploration, we now provide an overview of BAROM
 design. While many design methods envolved with new memory technologies endeavor
@@
 the bandwidth demand of a specific set of applications as the major metric. One
-desirable characteristic of BROM is that it provides prioritized access to the
+desirable characteristic of BAROM is that it provides prioritized access to the
 residual memory capacity available at runtime instead of sharing it among all
@@
 probablity-based predictor that can achieve higher accuracy with small
-performance overhead. The rest of this section describes the BROM design, which
+performance overhead. The rest of this section describes the BAROM design, which
 is composed of on-chip memory hierarchy hardware design, the reconfiguration
@@
 
-\subsection{Memory Hierarchy Design}
+\subsection{Hardware Configuration}
 
-hardware design: num of levels; memory technology of each level; size of each level
+Figure~\ref{fig:hw-overview} shows an overview of BAROM hardware configuration.
+The CMP platform consists of multiple cores, where L1 cahces are private to each
+core and lower level caches are shared by the cores. With the provided bandwidth
+vs. capacity curves of various memory technologies that are obtained using the
+method described in section~\ref{sec:tech-explore}, we can design a hybrid
+on-chip memory system to always provide the highest bandwith. In order to
+achieve this goal, we configure the memory hierarchy in the following
+manner.\vspace{5pt}
+
+\noindent{}{\bf{Number of levels.}} In order to decide the number of memory
+levels, we examine the provided bandwidth curve. Memory levels are seperated by
+the cross points of two memory technologies. For example,
+Figure~\ref{fig:memory-bw} illustrates that there are two cross points (SRAM and
+STT-RAM, STT-RAM and RRAM). Therefore, we will have three levels of shared
+caches.\vspace{5pt}
+
+\noindent{}{\bf{Memory technology of each level.}} The memory technology with
+the highest provided bandwidth within the given range of capacity of each level
+is selected. In our case, SRAM, STT-RAM, and RRAM are selected as the L2, L3, and
+L4 caches respectively. As a result, the overall provided bandwidth curve of the
+on-chip memory system will appear to be an envelope line of all the available
+memory technologies as shown in Figure~\ref{fig:pbw-overall}.\vspace{5pt}
+
+\noindent{}{\bf{Capacity of each level.}} The maximum availabe capacity of each
+level is determined by the cross point of two memory technologies. In our case,
+the capacities of SRAM/L2, STT-RAM/L2, and RRAM/L4 are 2MB, 512MB, and 1GB
+respectively.\vspace{5pt}
+
+When the system is initially powered on, the capacity of each level is defined
+to be the maximum available capacity of the level.
 
@@
 
-performance opt: lut long enough but not too long (num of entries, bits per entry)
+performance opt: lut long enough but not too long (num of entries, bits per
+entry)
 

### 2011-04-05 trunk/paper.tex

@@
 
-Figure~\ref{fig:hw-overview} shows an overview of BAROM hardware configuration.
-The CMP platform consists of multiple cores, where L1 cahces are private to each
+Figure~\ref{fig:hw-config} shows an overview of BAROM hardware configuration.
+The CMP system consists of multiple cores, where L1 cahces are private to each
 core and lower level caches are shared by the cores. With the provided bandwidth
@@
 the highest provided bandwidth within the given range of capacity of each level
-is selected. In our case, SRAM, STT-RAM, and RRAM are selected as the L2, L3, and
-L4 caches respectively. As a result, the overall provided bandwidth curve of the
-on-chip memory system will appear to be an envelope line of all the available
-memory technologies as shown in Figure~\ref{fig:pbw-overall}.\vspace{5pt}
+is selected. In our case, SRAM, STT-RAM, and RRAM are selected as the L2, L3,
+and L4 caches respectively. As a result, the overall provided bandwidth curve of
+the on-chip memory system will appear to be an envelope line of all the
+available memory technologies as shown in
+Figure~\ref{fig:pbw-envelope}.\vspace{5pt}
 
@@
 the capacities of SRAM/L2, STT-RAM/L2, and RRAM/L4 are 2MB, 512MB, and 1GB
-respectively.\vspace{5pt}
+respectively. At the system initialization, the capacity of each level is
+defined to be the maximum available capacity of the level. According to the
+reqirement of different applications, the capacities can be reconfigured at
+run-time to provide best performance. The memory space of each level can also be
+reconfigured to form an internal level as shown in Figure~\ref{fig:hw-config}.
+The hardware and software supports for reconfiguration will be presented in the
+following sections.
 
-When the system is initially powered on, the capacity of each level is defined
-to be the maximum available capacity of the level.
+\subsection{Reconfiguration}
+
+When the system is initially powered on,
 
@@
 modification to support reconfiguration; what to do with reconfigured capacity
-
-\subsection{Reconfiguration}
 

### 2011-04-05 trunk/paper.tex

@@
 is selected. In our case, SRAM, STT-RAM, and RRAM are selected as the L2, L3,
-and L4 caches respectively. As a result, the overall provided bandwidth curve of
-the on-chip memory system will appear to be an envelope line of all the
-available memory technologies as shown in
-Figure~\ref{fig:pbw-envelope}.\vspace{5pt}
+and L4 caches respectively. \vspace{5pt}
 
@@
 The hardware and software supports for reconfiguration will be presented in the
-following sections.
+following sections.\vspace{5pt}
+
+The goal of such architecture design is to make use of different memory
+technologies to configure an on-chip memory system with optimal provided
+bandwidth over a range of capacities. As a result, the overall provided
+bandwidth curve of the on-chip memory system will appear to be an envelope line
+of all the available memory technologies as illustrated in
+Figure~\ref{fig:pbw-envelope}. It is illustrated in this figure that the hybrid
+memory system always provides the highest bandwidth over different capacities.
 
@@
 
-When the system is initially powered on,
+Although the above hybrid memory system maintains the optimal provided bandwidth
+curve over all the capacities, it does not garantee the best performance for
+different applications with a variety of bandwidth requirements. The overall
+provided bandwidth curve shown in Figure~\ref{fig:pbs-envelope} appears to be
+monotonically decreasing. A specific application may have a bandwidth
+requirement that can be mapped to a single point on the curve. To address this
+issue, we dynamically reconfigure the memory space of each level corresponding
+to the bandwidth requirement of each set of applications running on the system.
+To be more aggressive, the reconfiguration is applied at the end of each given
+time interval to accomodate different program phases of an application.
 

### 2011-04-05 paper.tex

@@
 probablity-based predictor that can achieve higher accuracy with small
-performance overhead.
+performance overhead. The rest of this section describes the BAROM design, which
+is composed of on-chip memory hierarchy hardware design, the reconfiguration
+mechanism, and the predictor engine.
 
-\subsection{Memory Hierarchy Design}
+\subsection{Hardware Configuration}
+
+Figure~\ref{fig:hw-config} shows an overview of BAROM hardware configuration.
+The CMP system consists of multiple cores, where L1 cahces are private to each
+core and lower level caches are shared by the cores. With the provided bandwidth
+vs. capacity curves of various memory technologies that are obtained using the
+method described in section~\ref{sec:tech-explore}, we can design a hybrid
+on-chip memory system to always provide the highest bandwith. In order to
+achieve this goal, we configure the memory hierarchy in the following
+manner.\vspace{5pt}
+
+\noindent{}{\bf{Number of levels.}} In order to decide the number of memory
+levels, we examine the provided bandwidth curve. Memory levels are seperated by
+the cross points of two memory technologies. For example,
+Figure~\ref{fig:memory-bw} illustrates that there are two cross points (SRAM and
+STT-RAM, STT-RAM and RRAM). Therefore, we will have three levels of shared
+caches.\vspace{5pt}
+
+\noindent{}{\bf{Memory technology of each level.}} The memory technology with
+the highest provided bandwidth within the given range of capacity of each level
+is selected. In our case, SRAM, STT-RAM, and RRAM are selected as the L2, L3,
+and L4 caches respectively. \vspace{5pt}
+
+\noindent{}{\bf{Capacity of each level.}} The maximum availabe capacity of each
+level is determined by the cross point of two memory technologies. In our case,
+the capacities of SRAM/L2, STT-RAM/L2, and RRAM/L4 are 2MB, 512MB, and 1GB
+respectively. At the system initialization, the capacity of each level is
+defined to be the maximum available capacity of the level. According to the
+reqirement of different applications, the capacities can be reconfigured at
+run-time to provide best performance. The memory space of each level can also be
+reconfigured to form an internal level as shown in Figure~\ref{fig:hw-config}.
+The hardware and software supports for reconfiguration will be presented in the
+following sections.\vspace{5pt}
+
+The goal of such architecture design is to make use of different memory
+technologies to configure an on-chip memory system with optimal provided
+bandwidth over a range of capacities. As a result, the overall provided
+bandwidth curve of the on-chip memory system will appear to be an envelope line
+of all the available memory technologies as illustrated in
+Figure~\ref{fig:pbw-envelope}. It is illustrated in this figure that the hybrid
+memory system always provides the highest bandwidth over different capacities.
 
@@
 
-tailoring mode
-reuse mode
+Although the above hybrid memory configuration maintains the optimal provided
+bandwidth curve over all the capacities, it does not garantee the best
+performance for different applications with a variety of bandwidth requirements.
+The overall provided bandwidth curve shown in Figure~\ref{fig:pbs-envelope}
+appears to be monotonically decreasing. A specific application may have a
+bandwidth requirement that can be mapped to a single point on the curve. In
+order to satisfy the bandwidth demand of different applications, we dynamically
+reconfigure the memory space of each level corresponding to the bandwidth
+requirement of each set of applications running on the system. To be more
+aggressive, the reconfiguration is applied at the end of each given execution
+time interval to accomodate different program phases of applications.
+
+The major system parameter to be reconfigured is the capacity of each memory
+level. At each time interval, we determine the upper bound of the capacity of
+memory level-i ($s_{iu}$) by mapping the demand bandwidth (DBW) of a specific
+application to the system's provided bandwidth curve, i.e.,
+$s_{iu}=f^{-1}(\textrm{DBW})$ where $f(x)$ represents the provided bandwidth
+curve of the on-chip memory system stored in the look-up-table in a reserved
+non-volatile memory space. The demand bandwidth for each memory level is
+generated using the prediction engine presented in section~\ref{sec:predict}. In
+previous sections, we show that the memory with smaller capacity provides higher
+bandwidth. Theoritically, the higher bandwidth provided by the memory system
+leads to higher performance (in terms of throughput) and power consumption of
+throughput computing applications~\cite{throughput-vs-bandwidth}. Consequently,
+we define a lower bound of memory capacity as
+$s_{il}=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
+threshold to make sure possible provided bandwidth increase without much power
+overhead. The capacity of memory level-i ($s_i$) is therefore selected to
+satisfy $s_{il}\le s_i\le s_{iu}$.
+
+Figure~\ref{fig:reconfig} describes the hardware support for reconfiguration.
+
+reconfiguration support: how reconfigurable (Li -> Li+1 (Li+0.5); hardware
+modification to support reconfiguration.
+
+reconfigure the unused capacity:
+
+- if (> si): becomes si+1
+
+- else: unused (or becomes lut)
 
@@
 statistical prediction
-performance opt
+
+hierarchical prediction
+
+performance opt: lut long enough but not too long (num of entries, bits per
+entry)
 

### 2011-04-06 trunk/paper.tex

@@
 
-Although the above hybrid memory system maintains the optimal provided bandwidth
-curve over all the capacities, it does not garantee the best performance for
-different applications with a variety of bandwidth requirements. The overall
-provided bandwidth curve shown in Figure~\ref{fig:pbs-envelope} appears to be
-monotonically decreasing. A specific application may have a bandwidth
-requirement that can be mapped to a single point on the curve. To address this
-issue, we dynamically reconfigure the memory space of each level corresponding
-to the bandwidth requirement of each set of applications running on the system.
-To be more aggressive, the reconfiguration is applied at the end of each given
-time interval to accomodate different program phases of an application.
+Although the above hybrid memory configuration maintains the optimal provided
+bandwidth curve over all the capacities, it does not garantee the best
+performance for different applications with a variety of bandwidth requirements.
+The overall provided bandwidth curve shown in Figure~\ref{fig:pbs-envelope}
+appears to be monotonically decreasing. A specific application may have a
+bandwidth requirement that can be mapped to a single point on the curve. In
+order to satisfy the bandwidth demand of different applications, we dynamically
+reconfigure the memory space of each level corresponding to the bandwidth
+requirement of each set of applications running on the system. To be more
+aggressive, the reconfiguration is applied at the end of each given execution
+time interval to accomodate different program phases of applications.
+
+The major system parameter to be reconfigured is the capacity of each memory
+level. At each time interval, we determine the upper bound of the capacity of
+memory level-i ($s_{iu}$) by mapping the demand bandwidth (DBW) of a specific
+application to the system's provided bandwidth curve, i.e.,
+$s_{iu}=f^{-1}(\textrm{DBW})$ where $f(x)$ represents the provided bandwidth
+curve of the on-chip memory system stored in the look-up-table in a reserved
+non-volatile memory space. The demand bandwidth for each memory level is
+generated using the prediction engine presented in section~\ref{sec:predict}. In
+previous sections, we show that the memory with smaller capacity provides higher
+bandwidth. Theoritically, the higher bandwidth provided by the memory system
+leads to higher performance (in terms of throughput) and power consumption of
+throughput computing applications~\cite{throughput-vs-bandwidth}. Consequently,
+we define a lower bound of memory capacity as
+$s_{il}=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
+threshold to make sure possible provided bandwidth increase without much power
+overhead. The capacity of memory level-i ($s_i$) is therefore selected to
+satisfy $s_{il}\le s_i\le s_{iu}$.
+
+Our primary reconfigurable design exploits set associativity in conventional
+cache organizations. Figure~\ref{fig:reconfig} depicts the block diagram for the
+organization of our design. An n-way set associative cache consists of $n$ data
+and tag arrays. We divide the each level of on-chip memory into partitions at
+the granularity of the $k$-ways, where $k$ is determined by the number of
+available capacity values for a specific memory technology on the provided
+bandwidth curve. For example, if there are 8 available capacities for STT-RAM
+(from 2MB to 1GB), we will determine the granularity by $k=n/8$.The desirability
+of such configuration is to exploit the division of ways already present in a
+conventional cache organization.
+
+The index part of the input address is used to index all the
+ways of the data and tag array. The tag part of the input address is sent to the
+comparators of all the ways to determine if there is a match with any of the
+tags read from the tag array. If there is a match, a hit is signaled on the
+valid output line and data from the corresponding way of the data array is sent
+onto the output data lines.
 
 reconfiguration support: how reconfigurable (Li -> Li+1 (Li+0.5); hardware
-modification to support reconfiguration; what to do with reconfigured capacity
-
-decide size of each level si:
-
-1) theoritical curves
-
-2) bound:
-
-- dbwi < pbwi -> si <= s1
-
-- max throughput under power budget? -> si >= s2
-
-- si <= hardware given max si
-
-- si > si-1
+modification to support reconfiguration.
 

### 2011-04-06 paper.tex

@@
 
-\subsection{Hardware Configuration}
+\subsection{Overview}
 
@@
 the capacities of SRAM/L2, STT-RAM/L2, and RRAM/L4 are 2MB, 512MB, and 1GB
-respectively. At the system initialization, the capacity of each level is
-defined to be the maximum available capacity of the level. According to the
-reqirement of different applications, the capacities can be reconfigured at
-run-time to provide best performance. The memory space of each level can also be
-reconfigured to form an internal level as shown in Figure~\ref{fig:hw-config}.
-The hardware and software supports for reconfiguration will be presented in the
-following sections.\vspace{5pt}
+respectively. Each level of memory if further configured to be multiple banks
+the same way as the conventional cache design. At the system initialization, the
+capacity of each level is defined to be the maximum available capacity of the
+level. According to the reqirement of different applications, the capacities can
+be reconfigured at run-time to provide best performance. The memory space of
+each level can be reconfigured to form an internal level as shown in
+Figure~\ref{fig:hw-config}. The hardware and software supports for
+reconfiguration will be presented in the following sections.\vspace{5pt}
 
@@
 overhead. The capacity of memory level-i ($s_i$) is therefore selected to
-satisfy $s_{il}\le s_i\le s_{iu}$.
+satisfy $s_{il}\le s_i\le s_{iu}$. Furthermore, the rest of the capacity can be
+reconfigured to become level-i.5 if necessary. We describe the reconfiguration
+algorithm as follows.
 
-Figure~\ref{fig:reconfig} describes the hardware support for reconfiguration.
+<pseudocode>
 
-reconfiguration support: how reconfigurable (Li -> Li+1 (Li+0.5); hardware
-modification to support reconfiguration.
+Our primary reconfigurable design exploits set associativity in conventional
+cache organizations. The desirability of such configuration is to exploit the
+division of ways already present in a conventional cache organization.
+Figure~\ref{fig:reconfig} depicts the block diagram for the organization of our
+design. An n-way set associative cache consists of $n$ data and tag arrays. We
+divide the each level of on-chip memory into partitions at the granularity of
+the $k$-ways, where $k$ is determined by the available capacity range of a
+specific memory technology on the provided bandwidth curve. <100+ ways?
+experiment section> Reconfiguration will not affect the bits of the address
+fields that are used as tag, index, and block offset bits. Modifications to the
+conventional cache architecture include:
 
-reconfigure the unused capacity:
-
-- if (> si): becomes si+1
-
-- else: unused (or becomes lut)
+\begin{itemize}
+\item \textbf{Memory status vector} At each memory level, the corrent
+configuration is stored in a memory status vector.
+\item \textbf{Input and output paths} The input and output data paths are
+duplicated. 
+\item \textbf{Additional multiplexors} Futhermore, additional wiring and
+multiplexors at address decoders and tag comparators are also required to
+support the reconfiguration.
+\end{itemize}
 

### 2011-04-06 trunk/paper.tex

@@
 overhead. The capacity of memory level-i ($s_i$) is therefore selected to
-satisfy $s_{il}\le s_i\le s_{iu}$.
+satisfy $s_{il}\le s_i\le s_{iu}$. Furthermore, the rest of the capacity can be
+reconfigured to become level-i.5 if necessary. We describe the reconfiguration
+algorithm as follows.
+
+<pseudocode>
 
 Our primary reconfigurable design exploits set associativity in conventional
-cache organizations. Figure~\ref{fig:reconfig} depicts the block diagram for the
-organization of our design. An n-way set associative cache consists of $n$ data
-and tag arrays. We divide the each level of on-chip memory into partitions at
-the granularity of the $k$-ways, where $k$ is determined by the number of
-available capacity values for a specific memory technology on the provided
-bandwidth curve. For example, if there are 8 available capacities for STT-RAM
-(from 2MB to 1GB), we will determine the granularity by $k=n/8$.The desirability
-of such configuration is to exploit the division of ways already present in a
-conventional cache organization.
+cache organizations. The desirability of such configuration is to exploit the
+division of ways already present in a conventional cache organization.
+Figure~\ref{fig:reconfig} depicts the block diagram for the organization of our
+design. An n-way set associative cache consists of $n$ data and tag arrays. We
+divide the each level of on-chip memory into partitions at the granularity of
+the $k$-ways, where $k$ is determined by the available capacity range of a
+specific memory technology on the provided bandwidth curve. <100+ ways?
+experiment section> Reconfiguration will not affect the bits of the address
+fields that are used as tag, index, and block offset bits. Modifications to the
+conventional cache architecture include:
 
-The index part of the input address is used to index all the
-ways of the data and tag array. The tag part of the input address is sent to the
-comparators of all the ways to determine if there is a match with any of the
-tags read from the tag array. If there is a match, a hit is signaled on the
-valid output line and data from the corresponding way of the data array is sent
-onto the output data lines.
-
-reconfiguration support: how reconfigurable (Li -> Li+1 (Li+0.5); hardware
-modification to support reconfiguration.
-
-reconfigure the unused capacity:
-
-- if (> si): becomes si+1
-
-- else: unused (or becomes lut)
+\begin{itemize}
+\item \textbf{Memory status vector} At each memory level, the corrent
+configuration is stored in a memory status vector.
+\item \textbf{Input and output paths} The input and output data paths are
+duplicated. 
+\item \textbf{Additional multiplexors} Futhermore, additional wiring and
+multiplexors at address decoders and tag comparators are also required to
+support the reconfiguration.
+\end{itemize}
 
@@
 
+Memory reconfiguration relies on accurate predictions of demand bandwidth of a
+workload in response to the dynamically varying application characteristics.
+Conventional last value or table based predictors can model neither long range
+patterns of an application nor patterns with variable lengths. In this work, we
+employ a statistical predictor to support for the BAROM design. The basic idea
+of the predictor is borrowed from language modeling~\cite{smoothing-tech}.
+
+hardware: global predictor sitting before caches
 statistical prediction

### 2011-04-06 trunk/paper.tex

@@
 level. At each time interval, we determine the upper bound of the capacity of
-memory level-i ($s_{iu}$) by mapping the demand bandwidth (DBW) of a specific
+memory level-i ($s_i^u$) by mapping the demand bandwidth (DBW) of a specific
 application to the system's provided bandwidth curve, i.e.,
-$s_{iu}=f^{-1}(\textrm{DBW})$ where $f(x)$ represents the provided bandwidth
+$s_i^u=f^{-1}(\textrm{DBW})$ where $f(x)$ represents the provided bandwidth
 curve of the on-chip memory system stored in the look-up-table in a reserved
@@
 we define a lower bound of memory capacity as
-$s_{il}=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
+$s_i^l=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
 threshold to make sure possible provided bandwidth increase without much power
 overhead. The capacity of memory level-i ($s_i$) is therefore selected to
-satisfy $s_{il}\le s_i\le s_{iu}$. Furthermore, the rest of the capacity can be
+satisfy $s_i^l\le s_i\le s_i^u$. Furthermore, the rest of the capacity can be
 reconfigured to become level-i.5 if necessary. We describe the reconfiguration

### 2011-04-06 trunk/paper.tex

@@
 <pseudocode>
+if (si < stotali/2): 
+   if (si+1 <= stotali-si):
+      si.5 = stotali-si;
+
+else:
+   si+1 = ...;
 
@@
 employ a statistical predictor to support for the BAROM design. The basic idea
-of the predictor is borrowed from language modeling~\cite{smoothing-tech}.
+of the predictor is borrowed from the n-Gram models, which are typically used by
+speech and natural language processing~\cite{smoothing-tech}. The n-Gram models
+are usually formulated as a probability distribution $p(s)$ over a strings $s$,
+and attemp to reflect how frequently the string occurs as a sentense. Prediction
+on a reasonable combination of strings is then generated based on the
+probability distribution. Our predictor employs the same basic idea, but
+simplifies the model for performance considerations. In our model, each demand
+bandwidth sample abtained in a time interval is a string, and a given length
+(the ``order'') of samples is stored in a table as a pattern. We calculate the
+probability of each pattern at each time interval, and predict the demand
+bandwidth for the next time interval based on the probabilities.
+
+Figure~\ref{fig:predictor} shows the hardware components of our prediction
+engine, which includes a pattern look-up-table, a probability vector, and a set
+of counters.
 

### 2011-04-07 paper.tex

@@
   pure SRAM memory design. In addition, our design leads to \% throughput
-  increase and \% dynamic power reduction compared to hybrid but fixed memory
-  hierarchy design.
+  increase compared to hybrid but fixed memory hierarchy design.
 
@@
 \begin{itemize}
-\item Bandwidth-aware hybrid on-chip memory hierarchy design.
-\item Hardware support that enables prediction of cache performance on the
-  different sized caches.
-\item OS scheduler support to make use of the prediction capability and
-  appropriately schedule applications on to core with suitable cache capacity.
+\item Bandwidth-aware hybrid on-chip memory hierarchy design...
+\item Reconfiguration...
+\item Prediction engine...
 
@@
 reconfiguration. For each level of memory, we select the memory technology that
-has the highest provided bandwidth within given capacity. On top of such
-hardware hierarchy design, we apply dynamic memory system reconfiguration with
-the bandwidth demand of a specific set of applications as the major metric. One
-desirable characteristic of BAROM is that it provides prioritized access to the
-residual memory capacity available at runtime instead of sharing it among all
-data flows. By having control over the residual capacity, we can better utilize
-the limited resources to enhance system performance. The hardware
-reconfiguration is further facilitated by software prediction engine. Rather
-than conventional last value predictors and table-based predictors, we employ a
-probablity-based predictor that can achieve higher accuracy with small
-performance overhead. The rest of this section describes the BAROM design, which
-is composed of on-chip memory hierarchy hardware design, the reconfiguration
-mechanism, and the predictor engine.
+has the highest provided bandwidth within given capacity. The total capacity of
+each level is then partitioned to subsets of ways with different provided
+bandwidths. On top of such hardware hierarchy design, we apply dynamic memory
+system reconfiguration with the bandwidth demand of a specific set of
+applications as the major metric. One desirable characteristic of BAROM is that
+it provides prioritized access to the residual memory capacity available at
+runtime instead of sharing it among all memory accesses. By having control over
+the residual capacity, we can better utilize the limited resources to enhance
+system performance. The hardware reconfiguration is further facilitated by a
+prediction engine. Rather than conventional last value or history table based
+predictors, we employ a probablity-based statistical predictor which can achieve
+higher accuracy with small performance overhead. The rest of this section
+describes the BAROM design, which is composed of on-chip memory hierarchy
+design, the reconfiguration mechanism, and the prediction engine.
 
-\subsection{Overview}
+\subsection{Memory Hierarchy Design}
 
@@
 
+Memory reconfiguration relies on accurate predictions of demand bandwidth of a
+workload in response to the dynamically varying application characteristics.
+Conventional last value or table based predictors can model neither long range
+patterns of an application nor patterns with variable lengths. In this work, we
+employ a statistical predictor to support for the BAROM design. The basic idea
+of the predictor is borrowed from the n-Gram models, which are typically used by
+speech and natural language processing~\cite{smoothing-tech}. The n-Gram models
+are usually formulated as a probability distribution $p(s)$ over a strings $s$,
+and attemp to reflect how frequently the string occurs as a sentense. Prediction
+on a reasonable combination of strings is then generated based on the
+probability distribution. Our predictor employs the same basic idea, but
+simplifies the model for performance considerations. In our model, each demand
+bandwidth sample abtained in a time interval is a string, and a given length
+(the ``order'') of samples is stored in a table as a pattern. We calculate the
+probability of each pattern at each time interval, and predict the demand
+bandwidth for the next time interval based on the probabilities.
+
+Figure~\ref{fig:predictor} shows the hardware components of our prediction
+engine, which includes a pattern look-up-table, a probability vector, and a set
+of counters.
+
+The algorithm of the predictor is described in Figure~\ref{algo:predictor}.
+
+<pseudocode>
+
+hardware: global predictor sitting before caches
 statistical prediction
@@
 \section{Experiments}
+
+1. nvsim memory bandwidth vs. way-configuration -> reasonable fast ways and slow
+ways for reconfiguration
+
+2. predictor accuracy vs. benchmarks
+
+3. performance of instruction throughput vs. benchmarks
+
+------------------------------------
 

### 2011-04-07 trunk/paper.tex

@@
 of the predictor is borrowed from the n-Gram models, which are typically used by
-speech and natural language processing~\cite{smoothing-tech}. The n-Gram models
+speech and natural language processing~\cite{smoothing-tech}. The n-gram models
 are usually formulated as a probability distribution $p(s)$ over a strings $s$,
@@
 simplifies the model for performance considerations. In our model, each demand
-bandwidth sample abtained in a time interval is a string, and a given length
-(the ``order'') of samples is stored in a table as a pattern. We calculate the
-probability of each pattern at each time interval, and predict the demand
-bandwidth for the next time interval based on the probabilities.
+bandwidth (DBW) sample abtained in a time interval is analogous to a string in
+language, and a given length (the ``order'') of samples is stored in a table as
+a pattern. At each time interval, the prediction engine will update the pattern
+table with the new DBW sample, and calculate probability of the updated pattern.
+The probability of a pattern $s$ of length $l$ can be calculated without loss as
+the product of a sequence of conditional probabilities using the following
+equation.
+
+\begin{equation}
+  p(s)=p(w_1)p(w_2|w_1)...p(w_l|w_1...w_{l-1})=\prod_{i=1}^{l}p(w_i|w_1...w_{i-1})\label{eq:pr1}
+\end{equation}
+where $w_l$ is the DBW sample obtained in current time interval, and $w_1$
+through $w_{l-1}$ are the preceding $l-1$ DBW samples. In n-gram models, we make
+the approximation that each conditional probability only depends on the
+preceding $n-1$ samples and obtain Equation~\ref{eq:pr2}.
+
+\begin{equation}
+  p(s)=\prod_{i=1}^{l}p(w_i|w_{i-n+1}^{i-1})\label{eq:pr2}
+\end{equation}
+in which $w_{i-n+1}^{i-1}$ represents the sequence of $w_{i-n+1}...w_{i-1}$.
+
+The demand bandwidth in the next time interval is predicted based on the
+probabilities.
 
@@
 of counters.
+
+The algorithm of the predictor is described in Figure~\ref{algo:predictor}.
+
+<pseudocode>
 

### 2011-04-07 trunk/paper.tex

@@
 The probability of a pattern $s$ of length $l$ can be calculated without loss as
-the product of a sequence of conditional probabilities using the following
-equation.
+the product of a sequence of conditional probabilities using the equation shown
+below.
 
@@
 the approximation that each conditional probability only depends on the
-preceding $n-1$ samples and obtain Equation~\ref{eq:pr2}.
+preceding $n-1$ samples and obtain the following equation.
 
@@
 \end{equation}
-in which $w_{i-n+1}^{i-1}$ represents the sequence of $w_{i-n+1}...w_{i-1}$.
+in which $w_{i-n+1}^{i-1}$ denotes the sequence of $w_{i-n+1}...w_{i-1}$. In
+order to compute the result of Equation~\ref{eq:pr2}, an estimation of of
+$p(w_i|w_{i-n+1}^{i-1})$ can be generated using maximum likelihood~(ML)
+estimation method. In n-gram models, the largest n in wide use is n=3 which is
+called a trigram model. Therefore, we adopt n=3 in our predictor. Each
+conditional probability is calculated using ML estimation in
+Equation~\ref{eq:pr3}.
+
+\begin{equation}
+  \hat{p}(w_i|w_{i-n+1}^{i-1})=\frac{c(w_{i-2}^{i})}{c(w_{i-2}^{i-1})}\label{eq:pr3}
+\end{equation}
+where $c(w_{a}^{b})$ is the number of times that the sequence $w_a..w_b$ appears
+in preceding samples.
 

### 2011-04-09 paper.tex

@@
 performance computing systems to improve their computing performance by
-sacrifacing power efficiency. With the emerging 3D chip technology, caches can
+sacrificing power efficiency. With the emerging 3D chip technology, caches can
 be stacked on top of processor cores to provide high memory
@@
 Caching has long been employed as the most effective approach to reduce memory
-access latency. Adding an extra level of cache, however, can also help alleviate
-the increasing bandwidth pressure of off-chip memory. In [dac.1] the authors
-discuss extensively the problem of limited pin bandwidth to multiprocessor
-systems. They have focused on program performance in multiprocessor systems and
-make detailed decomposition of program execution times. They conclude that even
-more complex on-chip cache structures would prove to be costeffective, with the
-limited pin bandwidth severely restricting performance increases. In [dac.2] the
-authors have carefully studied the requirements for on-chip cache structures
-along with all sorts of optimization techniques that come along with scaling of
-processor core numbers. Their study shows that cache size need to grow much
-faster than processor core numbers to compensate for the limited off-chip
-bandwidth. Because of that, the near future processors need to allocate a huge
-percentage of chip area for caches, which means much less core counts than
-expected. The study also shows that effective bandwidth optimization techniques
-can help reduce cache size requirement and thus help scaling processor cores.
-However, the effectiveness of caching in the mid/large scale multiprocessor
-systems could be highly suboptimal due to significant contention across the
-parallel tasks. The more bandwidth requirement by a workload can be satisfied by
-the last level cache~(LLC), the less bandwidth will be required from the
-off-chip main memory. Our memory technology pre-exploration shows that the
-bandwidth provided by a memory decreases with the increase of its capacity.
-Based on these observations, adding more levels of cache can potentially fill in
-the bandwidth gap.
+access latency. Proper on-chip memory hierarchy design, however, can also help
+alleviate the increasing bandwidth pressure of off-chip memory. In [dac.1] the
+authors discuss extensively the problem of limited pin bandwidth to
+multiprocessor systems. They have focused on program performance in
+multiprocessor systems and make detailed decomposition of program execution
+times. They conclude that even more complex on-chip cache structures would prove
+to be costeffective, with the limited pin bandwidth severely restricting
+performance increases. In [dac.2] the authors have carefully studied the
+requirements for on-chip cache structures along with all sorts of optimization
+techniques that come along with scaling of processor core numbers. Their study
+shows that cache size need to grow much faster than processor core numbers to
+compensate for the limited off-chip bandwidth. Because of that, the near future
+processors need to allocate a huge percentage of chip area for caches, which
+means much less core counts than expected. The study also shows that effective
+bandwidth optimization techniques can help reduce cache size requirement and
+thus help scaling processor cores. However, the effectiveness of caching in the
+mid/large scale multiprocessor systems could be highly suboptimal due to
+significant contention across the parallel tasks. The more bandwidth requirement
+by a workload can be satisfied by the last level cache~(LLC), the less bandwidth
+will be required from the off-chip main memory. Our memory technology
+pre-exploration shows that the bandwidth provided by a memory decreases with the
+increase of its capacity. Based on these observations, adding more levels of
+cache can potentially fill in the bandwidth gap.
 
@@
 non-volatility, we demonstrate that these memory technologies provide higher
-bandwidth than SRAM at a high capacity. Consequently, it is beneficial to use
+bandwidth than SRAM at a large capacity. Consequently, it is beneficial to use
 NVM as lower level caches when large capacity.
@@
 requirement to off-chip main memory~\cite{Yu:2010:OCMB}. While most cache
-partitioning techniques focus on cache miss rates, our work takes a different
+partitioning techniques focus on cache miss rates, their work takes a different
 approach in which tasks¡¯ memory bandwidth requirements are taken into account
@@
 Based on our memory technology exploration, we now provide an overview of BAROM
-design. While many design methods envolved with new memory technologies endeavor
-to reduce the off-chip memory access latency, this work focuses on minimize
+design. While many design methods involved with new memory technologies endeavor
+to reduce the off-chip memory access latency, this work focuses on minimizing
 off-chip bandwidth requirement by employing hybrid on-chip memory and
@@
 prediction engine. Rather than conventional last value or history table based
-predictors, we employ a probablity-based statistical predictor which can achieve
-higher accuracy with small performance overhead. The rest of this section
-describes the BAROM design, which is composed of on-chip memory hierarchy
-design, the reconfiguration mechanism, and the prediction engine.
+predictors, we employ a probability-based statistical predictor which can
+achieve higher accuracy with small performance overhead. The rest of this
+section describes the BAROM design, which is composed of on-chip memory
+hierarchy design, the reconfiguration mechanism, and the prediction engine.
 
@@
 simplifies the model for performance considerations. In our model, each demand
-bandwidth sample abtained in a time interval is a string, and a given length
-(the ``order'') of samples is stored in a table as a pattern. We calculate the
-probability of each pattern at each time interval, and predict the demand
-bandwidth for the next time interval based on the probabilities.
+bandwidth (DBW) sample abtained in a time interval is analogous to a string in
+language, and a given length (the ``order'') of samples is stored in a table as
+a pattern. At each time interval, the prediction engine will update the pattern
+table with the new DBW sample, and calculate probability of the updated pattern.
+The probability of a pattern $s$ of length $l$ can be calculated without loss as
+the product of a sequence of conditional probabilities using the equation shown
+below.
+
+\begin{equation}
+  p(s)=p(w_1)p(w_2|w_1)...p(w_l|w_1...w_{l-1})=\prod_{i=1}^{l}p(w_i|w_1...w_{i-1})\label{eq:pr1}
+\end{equation}
+where $w_l$ is the DBW sample obtained in current time interval, and $w_1$
+through $w_{l-1}$ are the preceding $l-1$ DBW samples. In n-gram models, we make
+the approximation that each conditional probability only depends on the
+preceding $n-1$ samples and obtain the following equation.
+
+\begin{equation}
+  p(s)=\prod_{i=1}^{l}p(w_i|w_{i-n+1}^{i-1})\label{eq:pr2}
+\end{equation}
+in which $w_{i-n+1}^{i-1}$ denotes the sequence of $w_{i-n+1}...w_{i-1}$. In
+order to compute the result of Equation~\ref{eq:pr2}, an estimation of of
+$p(w_i|w_{i-n+1}^{i-1})$ can be generated using maximum likelihood~(ML)
+estimation method. In n-gram models, the largest n in wide use is n=3 which is
+called a trigram model. Therefore, we adopt n=3 in our predictor. Each
+conditional probability is calculated using ML estimation in
+Equation~\ref{eq:pr3}.
+
+\begin{equation}
+  \hat{p}(w_i|w_{i-n+1}^{i-1})=\frac{c(w_{i-2}^{i})}{c(w_{i-2}^{i-1})}\label{eq:pr3}
+\end{equation}
+where $c(w_{a}^{b})$ is the number of times that the sequence $w_a..w_b$ appears
+in preceding samples.
+
+The demand bandwidth in the next time interval is predicted based on the
+probabilities.
 

### 2011-04-09 paper.tex

@@
 bandwidth demand of the task, which ultimately determines the overall system
-performance and power consumption. <the moguls paper, the figure 1, however design for
-induvidual application, i.e., with different application can generate different
-designs.>
+performance and power consumption. However, their work only focused on LLC (L2
+cache). Our design targe is the overall on-chip memory system, which provides
+more design dimensions and flexibility. An analytical performance model was
+proposed by Sun et al. to explore the design space of memory hierarchies for
+throughput computing [moguls]. However, their memory hierarchy design was
+tailored for individual applications, which was undesirable for a real system
+design.
 
@@
 technologies are evaluated as LLCs by stacking each memory on top of the
-processor die. Both studies mainly focused on reducing the latency gap between L2
-cache~(LLC) and external memory and do not examine the bandwidth. <nvm cache
-research: most for latency, not bandwidth opt.>
+processor die. Both studies mainly focused on reducing the latency gap between
+L2 cache~(LLC) and external memory, but do not examine the design from bandwidth
+perspective.
 

### 2011-04-09 trunk/paper.tex

@@
 on a reasonable combination of strings is then generated based on the
-probability distribution. Our predictor employs the same basic idea, but
-simplifies the model for performance considerations. In our model, each demand
+probability distribution. Our predictor employs the same basic idea, but with
+very different implementation considerations. While language modeling is built
+from a set of previous collected training sentenses with finite lengths, our
+statistical model needs to be able to dynamically generate prediction based on a
+continuous sequence of metrics. In addition, we can only implement limited
+resolution for the metrics. Therefore, we need to normalize the demand bandwidth
+values with a limited number of quantization bins. In our model, each demand
 bandwidth (DBW) sample abtained in a time interval is analogous to a string in
@@
 where $c(w_{a}^{b})$ is the number of times that the sequence $w_a..w_b$ appears
-in preceding samples.
-
-The demand bandwidth in the next time interval is predicted based on the
-probabilities.
+in preceding samples. The demand bandwidth in the next time interval is
+predicted to be the last value in the pattern with the highest probability.
 
@@
 engine, which includes a pattern look-up-table, a probability vector, and a set
-of counters.
+of counters. The storage overhead
 
@@
 
-hardware: global predictor sitting before caches
-statistical prediction
-
-hierarchical prediction
-
-performance opt: lut long enough but not too long (num of entries, bits per
-entry)
+The computational complexity can be estimated as $O(ql)$, where $q$ is the
+number of quantization bins.
 

### 2011-04-09 trunk/paper.tex

@@
 the widening gap between the bandwidth demand created by processor cores and the
-limited access bandwidth provided by off-chip main memory [15][16][6][20]. Such
-limitation greatly affects the parallelism of memory demanding applications,
-i.e., applications with a large working set, by consuming additional cycles on
-off-chip memory access. In addition, even moderate memory demand applications
-will be affected as the number of cores scales up [17]. The continuing shrink of
+limited access bandwidth provided by off-chip main
+memory~\cite{McKee:2004:RMW,Burger:1996:MBLFM,Rogers:2009:SBW}. Such limitation greatly affects
+the parallelism of memory demanding applications, i.e., applications with a
+large working set, by consuming additional cycles on off-chip memory access. In
+addition, even moderate memory demand applications will be affected as the
+number of cores scales up~\cite{Huh:2001:EDSFC}. The continuing shrink of
 transistor density not only leads to increasingly powerful processors with a
@@
 power and cost efficient manner. High performance computing machines such as
-NVIDIA's Tesla~\cite{nvidia:tesla} rely on very high main memory bandwidths
-(provided by GDDR memories) to feed the requirement from the processor. However,
-GDDR memories run at higher clock rates and are more power hungry than
-conventional DRAM modules. It is undesirable for either general purpose or high
-performance computing systems to improve their computing performance by
-sacrificing power efficiency. With the emerging 3D chip technology, caches can
-be stacked on top of processor cores to provide high memory
-bandwidth[12][21][22]. However, ...
+NVIDIA's Tesla~\cite{Lindholm:2008:Tesla} rely on very high main memory
+bandwidths (provided by GDDR memories) to feed the requirement from the
+processor. However, GDDR memories run at higher clock rates and are more power
+hungry than conventional DRAM modules. It is undesirable for either general
+purpose or high performance computing systems to improve their computing
+performance by sacrificing power efficiency. With the emerging 3D chip
+technology, caches can be stacked on top of processor cores to provide high
+memory bandwidth~\cite{Sun:2009:E3DSL2}. However, ...
 
@@
 access latency. Proper on-chip memory hierarchy design, however, can also help
-alleviate the increasing bandwidth pressure of off-chip memory. In [dac.1] the
-authors discuss extensively the problem of limited pin bandwidth to
-multiprocessor systems. They have focused on program performance in
-multiprocessor systems and make detailed decomposition of program execution
-times. They conclude that even more complex on-chip cache structures would prove
-to be costeffective, with the limited pin bandwidth severely restricting
-performance increases. In [dac.2] the authors have carefully studied the
-requirements for on-chip cache structures along with all sorts of optimization
-techniques that come along with scaling of processor core numbers. Their study
-shows that cache size need to grow much faster than processor core numbers to
-compensate for the limited off-chip bandwidth. Because of that, the near future
-processors need to allocate a huge percentage of chip area for caches, which
-means much less core counts than expected. The study also shows that effective
-bandwidth optimization techniques can help reduce cache size requirement and
-thus help scaling processor cores. However, the effectiveness of caching in the
-mid/large scale multiprocessor systems could be highly suboptimal due to
-significant contention across the parallel tasks. The more bandwidth requirement
-by a workload can be satisfied by the last level cache~(LLC), the less bandwidth
-will be required from the off-chip main memory. Our memory technology
-pre-exploration shows that the bandwidth provided by a memory decreases with the
-increase of its capacity. Based on these observations, adding more levels of
-cache can potentially fill in the bandwidth gap.
+alleviate the increasing bandwidth pressure of off-chip memory. After an
+extensive study the problem of limited pin bandwidth to multiprocessor systems,
+Burger \emph{et al.} conclude that even more complex on-chip cache structures
+would prove to be costeffective with the limited pin bandwidth severely
+restricting performance increases~\cite{Burger:1996:MBLFM}. The requirements of
+on-chip cache structures along with all sorts of optimization techniques that
+come along with scaling of processor core numbers are carefully
+explored~\cite{Rogers:2009:SBW}. The study shows that cache size need to grow
+much faster than processor core numbers to compensate the limited off-chip
+bandwidth. The study also shows that effective bandwidth optimization techniques
+can help reduce cache size requirement and thus help scaling processor cores.
+Consequently, it is critical to re-design the on-chip memory system, which
+focuses on bandwidth optimization.
 
@@
 trade off between latency, bandwidth, and cost. Wu et al. explored various
-memory technologies - SRAM, eDRAM [23], magnetic RAM (MRAM) [12] and PCM to best
-construct L3 caches in terms of performance and power [24]. The eDRAM and MRAM
-technologies are evaluated as LLCs by stacking each memory on top of the
-processor die. Both studies mainly focused on reducing the latency gap between
-L2 cache~(LLC) and external memory, but do not examine the design from bandwidth
-perspective.
+memory technologies - SRAM, eDRAM~\cite{Wu:2009:HCA}, and
+STT-RAM~\cite{Sun:2009:NA3D} to best construct L3 caches in terms of performance
+and power. The eDRAM and STT-RAM technologies are evaluated as LLCs by stacking
+each memory on top of the processor die. Both studies mainly focused on reducing
+the latency gap between L2 cache~(LLC) and external memory, but do not examine
+the design from bandwidth perspective.
 
@@
 architectures. Zhou \emph{et al.} monitored memory access patterns and estimated
-memory behavior of workloads for energy efficient memory allocation
-[predictor.26]. Duesterwald et al. [10] describe different statistical and table
-based predictors for within- and across-metric predictions of performance
-monitoring information. They show that the table-based predictor generally
-outperforms the other predictors they tested. Sarikaya et al. [19] describe an
-optimal prediction technique based on a predictive least squares minimization.
-Sarikaya \emph{et al.} showed the benefit of statistical metric modeling for
-tracking varying pattern history lengths and modeling long term
-patterns~\cite{Sarikaya:2011:RWBP}. However, ...
+memory behavior of workloads for energy efficient memory
+allocation~\cite{Zhou:2004:DTP}. Duesterwald \emph{et al.} describe different
+statistical and table based predictors for within- and across-metric predictions
+of performance monitoring information~\cite{Duesterwald:2003:CPP}. They show
+that the table-based predictor generally outperforms the other predictors they
+tested. Sarikaya \emph{et al.}describe an optimal prediction technique based on
+a predictive least squares minimization~\cite{Sarikaya:2007:PPB}. Sarikaya
+\emph{et al.} showed the benefit of statistical metric modeling for tracking
+varying pattern history lengths and modeling long term
+patterns~\cite{Sarikaya:2010:RWB}. However, ...
 
@@
 % \includegraphics[width=2.5in]{figures/GST}
-% \caption{The schematic view of a PCRAM cell with MOSFET selector transistor (BL=Bitline, WL=Wordline, SL=Sourceline}
+% \caption{The schematic view of a PCRAM cell with MOSFET selector transistor
+% (BL=Bitline, WL=Wordline, SL=Sourceline}
 % \label{fig:GST}

### 2011-04-10 paper.tex

@@
 
-Figure~\ref{fig:hw-config} shows an overview of BAROM hardware configuration.
+\begin{figure}[t]
+\centering
+\includegraphics[width=2.5in]{figures/overview}
+\caption{Overview of BAROM hardware configuration.}
+\label{fig:overview}
+\end{figure}
+
+Figure~\ref{fig:overview} shows an overview of BAROM hardware configuration.
 The CMP system consists of multiple cores, where L1 cahces are private to each
@@
 each level can be reconfigured to form an internal level as shown in
-Figure~\ref{fig:hw-config}. The hardware and software supports for
+Figure~\ref{fig:overview}. The hardware and software supports for
 reconfiguration will be presented in the following sections.\vspace{5pt}

### 2011-04-10 paper.tex

@@
 
-Figure~\ref{fig:overview} shows an overview of BAROM hardware configuration.
+Figure~\ref{fig:overview} depicts an overview of BAROM hardware configuration.
 The CMP system consists of multiple cores, where L1 cahces are private to each
@@
 leads to higher performance (in terms of throughput) and power consumption of
-throughput computing applications~\cite{throughput-vs-bandwidth}. Consequently,
-we define a lower bound of memory capacity as
-$s_i^l=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
-threshold to make sure possible provided bandwidth increase without much power
-overhead. The capacity of memory level-i ($s_i$) is therefore selected to
-satisfy $s_i^l\le s_i\le s_i^u$. Furthermore, the rest of the capacity can be
-reconfigured to become level-i.5 if necessary. We describe the reconfiguration
-algorithm as follows.
+throughput computing applications. Consequently, we define a lower bound of
+memory capacity as $s_i^l=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a
+pre-defined threshold to make sure possible provided bandwidth increase without
+much power overhead. The capacity of memory level-i ($s_i$) is therefore
+selected to satisfy $s_i^l\le s_i\le s_i^u$. Furthermore, the rest of the
+capacity can be reconfigured to become level-i.5 if necessary. We describe the
+reconfiguration algorithm as follows.
 
@@
 division of ways already present in a conventional cache organization.
-Figure~\ref{fig:reconfig} depicts the block diagram for the organization of our
+Figure~\ref{fig:overview} shows the block diagram for the organization of our
 design. An n-way set associative cache consists of $n$ data and tag arrays. We
@@
 of the predictor is borrowed from the n-Gram models, which are typically used by
-speech and natural language processing~\cite{smoothing-tech}. The n-gram models
+speech and natural language processing~\cite{Chen:1996:ESS}. The n-gram models
 are usually formulated as a probability distribution $p(s)$ over a strings $s$,
@@
 engine, which includes a pattern look-up-table, a probability vector, and a set
-of counters. The storage overhead
+of counters. The storage overhead...
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=5in]{figures/predictor}
+\caption{Components of prediction engine.}
+\label{fig:predictor}
+\end{figure*}
 
